Improving the portuguese hyphenation rules

### 2023-07-19 article/article.tex
adding description of default TeX rules

\section{\TeX{} hyphenation rules for portuguese}\label{sec-tex-hyphen-pt}
\cite{rezende1987} was the first to create patterns for Portuguese hyphenation
in \TeX{}. The pattern set was last updated in 2015, incorporating
contributions from José João Dias Almeida \cite{hyphpt}, resulting in a total
of 307 rules\footnote{The set of rules for Portuguese hyphenation is short when
compared to rules in other languages. For example, English currently has 4938
rules, Russian has 7023 rules and German has 34011 rules.}. 
These rules effectively hyphenate the majority of Portuguese words.
% - detalhar as regras de Rezende e Dias 
% - contabilizar a quantidade de palavras que não são contempladas pela proposta do autor
% - analisar a necessidade de excluir alguma regra (ao invés de apenas acrescentar novas)

@@

>>>>

\section{\TeX{} hyphenation rules for portuguese}\label{sec-tex-hyphen-pt}
\cite{rezende1987} was the first to create patterns for Portuguese hyphenation
in \TeX{}. The pattern set was last updated in 2015, incorporating
contributions from José João Dias Almeida \cite{hyphpt}, resulting in a total
of 307 rules\footnote{The set of rules for Portuguese hyphenation is short when
compared to rules in other languages. For example, English currently has 4938
rules, Russian has 7023 rules and German has 34011 rules.}. 
These rules effectively hyphenate the majority of Portuguese words.

@@

%
% FILE="../data/default.TeX.pt-br.patterns"; grep "^1..$" "$FILE" | awk '{print substr($0, 0, 2)}' | sort | uniq -c | while read line; do p=$(echo $line | cut -d' ' -f2); c=$(grep "^$p[^0-9]$" "$FILE" | awk '{print substr($0, 3)}' | tr '\n' ' '); printf "$line $c \n"; done | cut -d' ' -f1 | paste -sd+ | bc
%
Out of the default \TeX{} rules, 252 follow the pattern \verb|1CV|, which
represents the recurring CV syllables in Portuguese.  As for the consonants,
there are typically 18 considered (\verb|b|, \verb|c|, \verb|ç|, \verb|d|,
\verb|f|, \verb|g|, \verb|j|, \verb|k|, \verb|l|, \verb|m|, \verb|n|, \verb|p|,
\verb|r|, \verb|s|, \verb|t|, \verb|v|, \verb|x|, and \verb|z|) that can
combine with 14 vowels (\verb|a|, \verb|e|, \verb|i|, \verb|o|, \verb|u|,
\verb|á|, \verb|â|, \verb|ã|, \verb|é|, \verb|í|, \verb|ó|, \verb|ú|, \verb|ê|,
and \verb|õ|), resulting in these CV patterns that indicate favorable
hyphenation points before the consonants.  It is worth noting that there might
be other rules or exceptions in the hyphenation patterns not covered by these
default rules.  To accommodate exceptions, 23 additional rules were created for
cases involving consonants \verb|b|, \verb|c|, \verb|d|, \verb|f|, \verb|g|,
\verb|k|, \verb|p|, \verb|t|, \verb|v|, or \verb|w| followed by \verb|l| or
\verb|r|; or \verb|c|, \verb|l| or \verb|n| followed by \verb|h|.

%
% FILE="../data/default.TeX.pt-br.patterns"; grep -v "^1..$" "$FILE" | grep "2" | awk '{print substr($0, 2, 1)}' | sort | uniq | while read line; doc=$(grep "^1${line}2.$" "$FILE" | tr -d '12' | awk '{print substr($0, 2, 1)}' | tr '\n' ' '); printf "$line $c\n"; done
%
Moreover, 23 distinct patterns were introduced to indicate hyphenation points
between vowels or between \verb|c|'s, \verb|r|'s, and \verb|s|'s.

%
% FILE="../data/default.TeX.pt-br.patterns"; grep -v "^1..$" "$FILE" | grep -v "2" | grep "3" | wc -l
%
Additionally, 8 patterns adhere to the \verb|1[gq]u4V|, signaling beneficial
hyphenation points before \verb|g| or \verb|q|, followed by a sequence of
\verb|u| and a vowel, with an inhibiting point between the \verb|u| and the
subsequent vowel.

Lastly, there is a single pattern represented as \verb|1-|, denoting that a
hyphen indeed serves as a beneficial hyphenation point."

% - contabilizar a quantidade de palavras que não são contempladas pela proposta do autor
% - analisar a necessidade de excluir alguma regra (ao invés de apenas acrescentar novas)
